---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# recurrentpseudo: An R package for analysing recurrent events in the presence of terminal events using pseudo-observations

<!-- badges: start -->
<!-- badges: end -->

This package computes pseudo-observations for recurrent event data in the presence of terminal events. Three versions exist: One-dimensional, two-dimensional or three-dimensional pseudo-observations. 


## Notation 
Let $D^*$ denote the survival time and let $N^*(t)$ denote the number of recurrent events by time $t$. 
Let $C$ denote the time of censoring. 
Due to right-censoring, the data consists of $X=\lbrace N(\cdot), D, \delta, Z \rbrace$ where $N(t) = N^*(t \wedge C)$, $D=D^* \wedge C$, $\delta = I \left( D^* \leq C \right)$ and $Z$ denotes $p$ baseline covariates.

We observe $X_i=\lbrace N_i(\cdot), D_i, \delta_i, Z_i \rbrace$ for each individual $i= 1,\ldots, n$.

We consider the marginal mean function, $\mu (t)$, given by
$$ 
\mu(t) = E(N^*(t)) = \int_0^t S(u^-) \, d R(u), \quad d R(t) = E(dN^*(t) \mid D^* \geq t)
$$
and the survival probability, $S(t)$ given by
$$ 
S(t) = P(D^*> t).
$$ 

Moreover, we consider the cumulative incidences for death causes 1, $C_1(t)$, and 2, $C_2(t)$
$$
C_1(t) = E(I(D^* \leq t, \Delta = 1)), \quad C_2(t) = E(I(D^* \leq t, \Delta = 2))
$$
where $\Delta = \lbrace 1, 2 \rbrace$ represents a cause-of-death indicator. 


## Introduction to pseudo-observations

The following section serves as a fast introduction to pseudo-observations, which the methods of this package is based on. 

For more detailed information, please see 
+ Andersen and Perme (_Pseudo-observations in survival analysis (2010)_) or 
+ Andersen, Klein and Rosthøj (_Generalised linear models for correlated pseudo-observations, with applications to multi-state models (2003)_)

We wish to formulate a model for
$$ 
\theta = E(f(X))
$$ 
where $X=X_1, \ldots, X_n$ denotes a vector of survival times (or other survival data) for $n$ individuals and $f$ denotes some function. 

An example would be $\theta = E(I(D^*>t)) = P(D^*>t)$. 

Assume that a sufficiently nice estimator $\hat{\theta}$ of $\theta$ exists. 

For a fixed time, $t \in [0, \tau]$, the pseudo-observation for the i'th individual at $t$ is given by 
$$ 
\hat{\theta}_i (t)= n \cdot \hat{\theta}(t) - (n-1) \cdot \hat{\theta}^{-i}(t)
$$
where $\hat{\theta}(t)$ denotes the estimate based on the total data set, and $\hat{\theta}^{-i}(t)$ denotes the estimate based on the same data set but omitting observations from individual i. 

Since the survival times are subject to right-censoring, inference based on likelihood is adjusted to accommodate this. 


However, since all subjects has a valid pseudo-observation, $\hat{\theta}_i (t)$, at one or more times, these can be used as an outcome variable in a generalised linear model. 

Note, that this is regardless of the whether a subject is censored or died at time t. 

Assume that $g$ denotes a link function, then we wish to fit

$$
g(E(f(X) \mid Z)) = \xi^T Z.
$$
Following, $f(X)$ is replaced by $\hat{\theta}_i (\cdot)$ in the model fit. 


The model parameters, $\xi$, are estimated using generalised estimating equations, see Liang and Zeger (_Longitudinal data analysis using generalized linear models (1986)_. 

This can accomodate the fact that each individual can have several pseudo-observations if more than one time point is used in the computation. 




## One-dimensional pseudo-observations 
The one-dimensional pseudo-observations model is based on the parameter
$$
\hat{\theta} = \hat{\mu}(t) =  \int_0^t \hat{S}(u^-) \, d \hat{R}(u),
$$ 
where $\hat{S}(t)$ denotes the Kaplan-Meier estimator of $S(t)$ and $\hat{R}(t)$ denotes the Nelson-Aalen estimator of $R(t)$.


## Two-dimensional pseudo-observations 
The two-dimensional pseudo-observations model is based on the parameter
$$
\hat{\theta} = \left( \begin{matrix} \hat{\mu}(t) \\ \hat{S}(t) \end{matrix} \right)
$$ 


## Three-dimensional pseudo-observations 
The three-dimensional pseudo-observations model is based on the parameter
$$
\hat{\theta} = \left( \begin{matrix} \hat{\mu}(t) \\ \hat{C}_1(t) \\ \hat{C}_2(t) \end{matrix} \right)
$$ 
where $\hat{C}_1(t)$ and $\hat{C}_2(t)$ are the Aalen-Johansen estimates of the cumulative incidences for causes 1, $C_1(t)$, and 2, $C_2(t)$, respectively.



# Install package from GitHub


```{r eval = T, echo = FALSE}
require(devtools)

devtools::install_github("JulieKFurberg/recurrentpseudo", force = TRUE)

require(recurrentpseudo)
```


# Example - Bladder cancer data from survival package


We consider the well-known bladder cancer data from the survival package.

We focus on the comparison between placebo and thiotepa. 

```{r cache = F}
# Example: Bladder cancer data from survival package
require(survival)

# Make a three level status variable
bladder1$status3 <- ifelse(bladder1$status %in% c(2, 3), 2, bladder1$status)

# Add one extra day for the two patients with start=stop=0
# subset(bladder1, stop <= start)
bladder1[bladder1$id == 1, "stop"] <- 1
bladder1[bladder1$id == 49, "stop"] <- 1

# Restrict the data to placebo and thiotepa
bladdersub <- subset(bladder1, treatment %in% c("placebo", "thiotepa"))

# Make treatment variable two-level factor
bladdersub$Z <- as.factor(ifelse(bladdersub$treatment == "placebo", 0, 1))
levels(bladdersub$Z) <- c("placebo", "thiotepa")
head(bladdersub)

```


One-dimensional pseudo-observations can be computed using the following code

```{r cache = F}
# Pseudo observations
pseudo_bladder_1d <- pseudo.onedim(tstart = bladdersub$start,
                                   tstop = bladdersub$stop,
                                   status = bladdersub$status3,
                                   id = bladdersub$id,
                                   covar_names = "Z",
                                   tk = c(20, 30, 40),
                                   data = bladdersub)
head(pseudo_bladder_1d$outdata)

# GEE fit
fit_bladder_1d <- pseudo.geefit(pseudodata = pseudo_bladder_1d,
                                covar_names = c("Z"))
fit_bladder_1d

# Treatment differences
xi_diff_1d <- as.matrix(c(fit_bladder_1d$xi[4]), ncol = 1)

mslabels <- c("treat, mu")
rownames(xi_diff_1d) <- mslabels
colnames(xi_diff_1d) <- ""
xi_diff_1d

# Variance matrix for differences
sigma_diff_1d <- matrix(c(fit_bladder_1d$sigma[4,4]),
                          ncol = 1, nrow = 1,
                          byrow = T)

rownames(sigma_diff_1d) <- colnames(sigma_diff_1d) <- mslabels
sigma_diff_1d
```



Two-dimensional pseudo-observations can be computed using the following code

```{r cache = F}
# Pseudo observations
pseudo_bladder_2d <- pseudo.twodim(tstart = bladdersub$start,
                                   tstop = bladdersub$stop,
                                   status = bladdersub$status3,
                                   id = bladdersub$id,
                                   covar_names = "Z",
                                   tk = c(20, 30, 40),
                                   data = bladdersub)
head(pseudo_bladder_2d$outdata)

# GEE fit
fit_bladder_2d <- pseudo.geefit(pseudodata = pseudo_bladder_2d,
                                covar_names = c("Z"))
fit_bladder_2d

# Treatment differences
xi_diff_2d <- as.matrix(c(fit_bladder_2d$xi[4] - fit_bladder_2d$xi[1],
                          fit_bladder_2d$xi[8] - fit_bladder_2d$xi[6]), ncol = 1)

mslabels <- c("treat, mu", "treat, surv")
rownames(xi_diff_2d) <- mslabels
colnames(xi_diff_2d) <- ""
xi_diff_2d


# Variance matrix for differences
sigma_diff_2d <- matrix(c(fit_bladder_2d$sigma[4,4] + fit_bladder_2d$sigma[1,1],
                          fit_bladder_2d$sigma[1,4] + fit_bladder_2d$sigma[1,5],
                          fit_bladder_2d$sigma[5,8] + fit_bladder_2d$sigma[5,4],
                          fit_bladder_2d$sigma[8,8] + fit_bladder_2d$sigma[5,5]),
                          ncol = 2, nrow = 2,
                          byrow = T)

rownames(sigma_diff_2d) <- colnames(sigma_diff_2d) <- mslabels
sigma_diff_2d
```

Three-dimensional pseudo-observations can be computed using the following code

```{r eval = T}
# Add deathtype variable to bladder data
# Deathtype = 1 (bladder disease death), deathtype = 2 (other death reason)
bladdersub$deathtype <- with(bladdersub, ifelse(status == 2, 1, ifelse(status == 3, 2, 0)))
table(bladdersub$deathtype, bladdersub$status)

# Pseudo-observations
pseudo_bladder_3d <- pseudo.threedim(tstart = bladdersub$start,
                                     tstop = bladdersub$stop,
                                     status = bladdersub$status3,
                                     id = bladdersub$id,
                                     deathtype = bladdersub$deathtype,
                                     covar_names = "Z",
                                     tk = c(20, 30, 40),
                                     data = bladdersub)
head(pseudo_bladder_3d$outdata_long)

# GEE fit
fit_bladder_3d <- pseudo.geefit(pseudodata = pseudo_bladder_3d,
                                covar_names = c("Z"))
fit_bladder_3d

# Treatment differences
xi_diff_3d <- as.matrix(c(fit_bladder_3d$xi[4] - fit_bladder_3d$xi[1],
                          fit_bladder_3d$xi[8] - fit_bladder_3d$xi[5],
                          fit_bladder_3d$xi[12] - fit_bladder_3d$xi[9]), ncol = 1)

mslabels <- c("treat, mu", "treat, cif1", "treat, cif2")
rownames(xi_diff_3d) <- mslabels
colnames(xi_diff_3d) <- ""
xi_diff_3d


# Variance matrix for differences
sigma_diff_3d <- matrix(c(fit_bladder_3d$sigma[1,1] + fit_bladder_3d$sigma[4,4],
                          fit_bladder_3d$sigma[1,9] + fit_bladder_3d$sigma[1,5],
                          fit_bladder_3d$sigma[1,8] + fit_bladder_3d$sigma[1,12],

                          fit_bladder_3d$sigma[1,9] + fit_bladder_3d$sigma[1,5],
                          fit_bladder_3d$sigma[5,5] + fit_bladder_3d$sigma[8,8],
                          fit_bladder_3d$sigma[9,5] + fit_bladder_3d$sigma[8,12],

                          fit_bladder_3d$sigma[1,8] + fit_bladder_3d$sigma[1,12],
                          fit_bladder_3d$sigma[9,5] + fit_bladder_3d$sigma[8,12],
                          fit_bladder_3d$sigma[9,9] + fit_bladder_3d$sigma[12,12]

                          ),
                        ncol = 3, nrow = 3,
                        byrow = T)

rownames(sigma_diff_3d) <- colnames(sigma_diff_3d) <- mslabels
sigma_diff_3d

```


```{r eval = T}
###----------------------- Compare - should match for mu elements ---------------------------###
xi_diff_1d
xi_diff_2d
xi_diff_3d

sigma_diff_1d
sigma_diff_2d
sigma_diff_3d
```


```{r eval = T}
# if more than a binary treat

## One-dim
# Treatment, binary variable:
pseudo_bladder_1d$outdata_long$Z1_ <- as.factor(pseudo_bladder_1d$outdata_long$Z)

# A continuous variable
pseudo_bladder_1d$outdata_long$Z2_ <- rnorm(nrow(pseudo_bladder_1d$outdata_long), mean = 3, sd = 1)

# A categorical variable
pseudo_bladder_1d$outdata_long$Z3_ <- sample(x = c("A", "B", "C"), 
                                 size = nrow(pseudo_bladder_1d$outdata_long), 
                                 replace = TRUE, 
                                 prob = c(1/4, 1/2, 1/4))

fit1 <- pseudo.geefit(pseudodata = pseudo_bladder_1d, 
                      covar_names = c("Z1_", "Z2_", "Z3_"))

fit1$xi
fit1$sigma

## Two-dim
# Treatment, binary variable:
pseudo_bladder_2d$outdata_long$Z1_ <- as.factor(pseudo_bladder_2d$outdata_long$Z)

# A continuous variable
pseudo_bladder_2d$outdata_long$Z2_ <- rnorm(nrow(pseudo_bladder_2d$outdata_long), mean = 3, sd = 1)

# A categorical variable
pseudo_bladder_2d$outdata_long$Z3_ <- sample(x = c("A", "B", "C"), 
                                 size = nrow(pseudo_bladder_2d$outdata_long), 
                                 replace = TRUE, 
                                 prob = c(1/4, 1/2, 1/4))

fit2 <- pseudo.geefit(pseudodata = pseudo_bladder_2d, 
                      covar_names = c("Z1_", "Z2_", "Z3_"))

fit2$xi
fit2$sigma

## Three-dim
# Treatment, binary variable:
pseudo_bladder_3d$outdata_long$Z1_ <- as.factor(pseudo_bladder_3d$outdata_long$Z)

# A continuous variable
pseudo_bladder_3d$outdata_long$Z2_ <- rnorm(nrow(pseudo_bladder_3d$outdata_long), mean = 3, sd = 1)

# A categorical variable
pseudo_bladder_3d$outdata_long$Z3_ <- sample(x = c("A", "B", "C"), 
                                 size = nrow(pseudo_bladder_3d$outdata_long), 
                                 replace = TRUE, 
                                 prob = c(1/4, 1/2, 1/4))

fit3 <- pseudo.geefit(pseudodata = pseudo_bladder_3d, 
                      covar_names = c("Z1_", "Z2_", "Z3_"))

fit3$xi
fit3$sigma

```

# Citation 
To cite the `recurrentpseudo` package please use the following references,

> Julie K. Furberg, Per K. Andersen, Sofie Korn, Morten Overgaard, Henrik Ravn: Bivariate pseudo-observations for recurrent event analysis with terminal events (Lifetime Data Analysis, 2021)

