---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# recurrentpseudo: An R package for analysing recurrent events in the presence of terminal events using pseudo-observations

<!-- badges: start -->
<!-- badges: end -->

This package computes pseudo-observations for recurrent event data in the presence of terminal events. Three versions exist: One-dimensional, two-dimensional or three-dimensional pseudo-observations. 


## Notation 
Let $D^*$ denote the survival time and let $N^*(t)$ denote the number of recurrent events by time $t$. 
Let $C$ denote the time of censoring. 
Due to right-censoring, the data consists of $X=\lbrace N(\cdot), D, \delta, Z \rbrace$ where $N(t) = N^*(t \wedge C)$, $D=D^* \wedge C$, $\delta = I \left( D^* \leq C \right)$ and $Z$ denotes $p$ baseline covariates.

We observe $X_i=\lbrace N_i(\cdot), D_i, \delta_i, Z_i \rbrace$ for each individual $i= 1,\ldots, n$.

We consider the marginal mean function, $\mu (t)$, given by
$$ 
\mu(t) = E(N^*(t)) = \int_0^t S(u^-) \, d R(u), \quad d R(t) = E(dN^*(t) \mid D^* \geq t)
$$
and the survival probability, $S(t)$ given by
$$ 
S(t) = P(D^*> t).
$$ 

Moreover, we consider the cumulative incidences for death causes 1, $C_1(t)$, and 2, $C_2(t)$
$$
C_1(t) = E(I(D^* \leq t, \Delta = 1)), \quad C_2(t) = E(I(D^* \leq t, \Delta = 2))
$$
where $\Delta = \lbrace 1, 2 \rbrace$ represents a cause-of-death indicator. 


## Introduction to pseudo-observations

## One-dimensional pseudo-observations 
The one-dimensional pseudo-observations model is based on the parameter
$$
\hat{\theta} = \hat{\mu}(t) =  \int_0^t \hat{S}(u^-) \, d \hat{R}(u),
$$ 
where $\hat{S}(t)$ denotes the Kaplan-Meier estimator of $S(t)$ and $\hat{R}(t)$ denotes the Nelson-Aalen estimator of $R(t)$.


## Two-dimensional pseudo-observations 
The two-dimensional pseudo-observations model is based on the parameter
$$
\hat{\theta} = \left( \begin{matrix} \hat{\mu}(t) \\ \hat{S}(t) \end{matrix} \right)
$$ 


## Three-dimensional pseudo-observations 
The three-dimensional pseudo-observations model is based on the parameter
$$
\hat{\theta} = \left( \begin{matrix} \hat{\mu}(t) \\ \hat{C}_1(t) \\ \hat{C}_2(t) \end{matrix} \right)
$$ 
where $\hat{C}_1(t)$ and $\hat{C}_2(t)$ are the Aalen-Johansen estimates of the cumulative incidences for causes 1, $C_1(t)$, and 2, $C_2(t)$, respectively.



# Install package from GitHub


```{r eval = FALSE}
devtools::install_github("JulieKFurberg/recurrentpseudo")
```


# Citation 
To cite the `recurrentpseudo` package please use the following references,

> Julie K. Furberg, Per K. Andersen, Sofie Korn, Morten Overgaard, Henrik Ravn: Bivariate pseudo-observations for recurrent event analysis with terminal events (Lifetime Data Analysis, 2021)

